Parameter Inference\\with Bifurcation Diagrams

### 2021-05-23 docs/article.tex
Updates from Overleaf

\begin{itemize}
    \item An end-to-end differentiable method for searching for bifurcations and then matching them to user-specified target locations
    \item Implementation of the method in Julia package leveraging automatic differentiation tooling
    \item Leveraging the cost landscape for a novel way of organising differential equation models in terms of geometric and topological equivalence
\end{itemize}

@@

\subsection{Genetic Toggle Switch}
In this section we optimise a model where the states share a Hill function relationship with cooperatively $n=2$; these models often emerge from mass action kinetics with quasi-steady state approximations and are used to model species concentrations. After re-scaling the equations governing the dynamics of concentrations, the simplified equations for state $u_1$ and $u_2$ become 
\begin{equation}
    \partial_t u_1 = \frac{ a_1 + (p u_2)^2}{ 1 + (p u_2)^2 } - \mu_1 u_1 \quad
    \partial_t u_2 = \frac{ a_2 + (k u_1)^2}{ 1 + (k u_1)^2 } - \mu_2 u_2
    \label{eq:two-state}
\end{equation}
where $a_k$ is the baseline production rate for species $k$ in the absence of the other species. Each species has a finite degradation rate $\mu_k$. Finally we have two half-saturation constants $p$ and $k$, one of which is chosen as our control condition. A baseline production rate $a_k>1$ recovers an inhibitor type hill function for species $k$ and is an activator otherwise. The half-saturation are proportional to the slope of the hill activations.

%  the  we can recover both activator and inhibitor relationships.
% Figure \ref{fig:two-state-optima} shows a UMAP embedding of the optimal parameter basins, represented as point clouds. 16 distinct clusters are obtained using DBSCAN. Each cluster corresponds to a kinetically different model regions that are qualitatively equivalent.

\begin{figure}
\centering
\includegraphics[width=11cm]{two-state-optima}
\caption{Optimal parameter estimates $\theta^*$ for the two state model \eqref{eq:two-state} for the targets $\targets=\{4,5\}$ reveal two clusters of qualitatively different regimes: mutual activation and mutual inhibition. }
\label{fig:two-state-optima}
\end{figure}

>>>>

\begin{itemize}
    \item An end-to-end differentiable method for searching for bifurcations and then matching them to user-specified target locations
    \item Implementation of the method as a Julia package \cite{Szep2021BifurcationFit.jlDiagrams}
    \item Leveraging the cost landscape for a novel way of organising differential equation models in terms of geometric and topological equivalence
\end{itemize}

@@

\subsection{Genetic Toggle Switch}
In this section we optimise a model where the states share a Hill function relationship with cooperatively $n=2$; these models often emerge from mass action kinetics with quasi-steady state approximations and are used to model species concentrations. After re-scaling the equations governing the dynamics of concentrations, the simplified equations for state $u_1$ and $u_2$ become 
\begin{equation}
    \partial_t u_1 = \frac{ a_1 + (p u_2)^2}{ 1 + (p u_2)^2 } - \mu_1 u_1 \quad
    \partial_t u_2 = \frac{ a_2 + (k u_1)^2}{ 1 + (k u_1)^2 } - \mu_2 u_2
    \label{eq:two-state}
\end{equation}
where $a_k$ is the baseline production rate for species $k$ in the absence of the other species. Each species has a finite degradation rate $\mu_k$. Finally we have two half-saturation constants $p$ and $k$, one of which is chosen as our control condition. A baseline production rate $a_k>1$ recovers an inhibitor type hill function for species $k$ and is an activator otherwise. The half-saturation are proportional to the slope of the hill activations. Solving for the steady states and substituting the equation for $u_1$ into $u_2$ gives rise to
\begin{equation}
    \mu_2 u_2 = \frac{ a_2 + (\frac{k}{\mu_1}
    \frac{ a_1 + (p u_2)^2}{ 1 + (p u_2)^2 })^2}{ 1 + (\frac{k}{\mu_1} 
    \frac{ a_1 + (p u_2)^2}{ 1 + (p u_2)^2 }
    )^2 }
    \label{eq:steady-states}
\end{equation}
which reveals that only the ratio between half-saturation and degradation $\frac{k}{\mu_1}$ affects the solutions to this equation, and hence the locations of the bifurcations. Figure \ref{fig:minimal-models:results} reveals the results for 800 optimisation runs with 200 ADAM iterations. 98\% of runs converged onto one of the two clusters: mutual activation and mutual inhibition regimes, examples of which are shown as insets.
\begin{figure}
\centering
\includegraphics[width=11cm]{two-state-optima}
\caption{Optimal parameter estimates $\theta^*$ for the two state model \eqref{eq:two-state} for the targets $\targets=\{4,5\}$ reveal two clusters of qualitatively different regimes: mutual activation and mutual inhibition. }
\label{fig:two-state-optima}
\end{figure}
\subsection{Complexity}

\begin{figure}
\centering
\includegraphics[width=11cm]{scaling}
\caption{Complexity scaling of calculating the gradient of the cost function}
\label{fig:scaling}
\end{figure}
